---
title: "lab12"
output:
  pdf_document: default
  html_document: default
date: "2024-04-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
8.1


Wide Format Example in Base R


Imagine we have a dataset of three subjects who have been measured at three different time points. In a wide format, each subject would be a row, and each measurement time point would be a column.
```{r}
# Creating a wide format data frame
subjects <- c(1, 2, 3)
time1 <- c(5, 4, 7)
time2 <- c(6, 5, 8)
time3 <- c(7, 6, 9)

wide_data <- data.frame(subjects, time1, time2, time3)
print(wide_data)
```
Converting from Wide to Long Format in Base R


To convert from wide to long format, we can use the following:
```{r}
# Convert the wide data frame to long format
long_data <- reshape(wide_data, 
                     varying = c("time1", "time2", "time3"), 
                     v.names = "measurement",
                     timevar = "time",
                     idvar = "subjects",
                     direction = "long")
print(long_data)
```
In this reshape function:
varying: Lists the columns in wide_data that vary by time point.
v.names: Names the measurement variable in the long format.
timevar: Names the variable that will record the time point in the long format.
idvar: Identifies the variable that identifies each subject.
direction: Specifies the direction of the reshape, "long" for wide to long.


The resulting long_data will have one row for each subject and time point, with the measurements in a single column.


8.4: Dropout Mechanisms

1. MCAR (Missing Completely at Random)
```{r}
set.seed(123) # for reproducibility
data_mcar <- data.frame(
  x = rnorm(100), # some continuous variables
  y = rnorm(100)
)
# Introduce missing values completely at random
data_mcar$x[sample(1:100, 10)] <- NA
data_mcar$y[sample(1:100, 10)] <- NA
```


2. MAR (Missing at Random)
```{r}
set.seed(123)
data_mar <- data.frame(
  x = rnorm(100),
  y = rnorm(100),
  group = sample(LETTERS[1:2], 100, replace = TRUE) # categorical group
)
# Introduce missing values depending on the group (but not on Y itself)
data_mar$x[data_mar$group == "A"] <- ifelse(runif(sum(data_mar$group == "A")) < 0.2, NA, data_mar$x[data_mar$group == "A"])
data_mar$y[data_mar$group == "B"] <- ifelse(runif(sum(data_mar$group == "B")) < 0.2, NA, data_mar$y[data_mar$group == "B"])
```

3.3. MNAR (Missing Not at Random or Non-Ignorable)
```{r}
set.seed(123)
data_mnar <- data.frame(
  x = rnorm(100),
  y = rbinom(100, 1, 0.5) # some binary variable
)
# Introduce missing values in x depending on the value of y
data_mnar$x[data_mnar$y == 1] <- ifelse(runif(sum(data_mnar$y == 1)) < 0.3, NA, data_mnar$x[data_mnar$y == 1])
```